Certified Knowledge Compilation \\ with Application to Verified Model Counting

### 2023-03-21 doc/lean.tex
save some lines in Lean

The invariants refer to the checker state \lstinline{st} with fields \lstinline{st.inputCnf} for $\inputformula$, \lstinline{st.clauseDb} for $\theta$, \lstinline{st.pog} for $P$, \lstinline{st.pogDefsForm} for the POG definitions formula $\bigwedge_{\nodeu\in P}\theta_u$, and \lstinline{st.allVars} for all variables (original and extension) added so far. For any $\nodeu\in P$, \lstinline{st.pog.toPropForm u} computes $\phi_\nodeu$. The first two invariants state that assignments to original variables extend uniquely to extension variables defining the POG nodes. In the formalization, we split this into two parts, namely, extension and uniqueness:
\begin{lstlisting}
/-- Any assignment satisfying φ₁ extends to φ₂ while preserving values on X. -/
def extendsOver (X : Set Var) (φ₁ φ₂ : PropForm Var) :=
  ∀ (σ₁ : PropAssignment Var), σ₁ ⊨ φ₁ → ∃ σ₂, σ₁.agreeOn X σ₂ ∧ σ₂ ⊨ φ₂

invariants.extends_pogDefsForm : extendsOver st.inputCnf.vars ⊤ st.pogDefsForm

@@

invariants.uep_pogDefsForm : uniqueExt st.inputCnf.vars st.allVars st.pogDefsForm
\end{lstlisting}
The next one guarantees that the set of solutions over the original variables is preserved:
\begin{lstlisting}
def equivalentOver (X : Set Var) (φ₁ φ₂ : PropForm Var) :=
  extendsOver X φ₁ φ₂ ∧ extendsOver X φ₂ φ₁

invariants.equivalent_clauseDb : equivalentOver st.inputCnf.vars
  st.inputCnf st.clauseDb
\end{lstlisting}
Finally, for every node $\nodeu\in P$ with corresponding literal $u$ we ensure that $\phi_\nodeu$ is partitioned (\cref{def:partitioned-operation-formula}) and relate $\phi_\nodeu$ to its clausal encoding $\theta_u \doteq u \wedge \bigwedge_{\nodev\in P}\theta_v$:
\begin{lstlisting}
def partitioned : PropForm Var → Prop
  | tr | fls | var _ => True
  | neg φ    => φ.partitioned
  | disj φ ψ => φ.partitioned ∧ ψ.partitioned ∧ ∀ τ, ¬(τ ⊨ φ ∧ τ ⊨ ψ)
  | conj φ ψ => φ.partitioned ∧ ψ.partitioned ∧ φ.vars ∩ ψ.vars = ∅

invariants.partitioned : ∀ u : ILit, (st.pog.toPropForm u).partitioned

invariants.equivalent_lits : ∀ u : ILit, equivalentOver st.inputCnf.vars
    (u ∧ st.pogDefsForm) (st.pog.toPropForm x)
\end{lstlisting}
These are maintained by valid CRAT proofs. Together with a few additional invariants
that ensure the correctness of cached computations, they imply the soundness theorem for $P$ with root
node $\noder$: the \textsc{final conditions} ensure that $\theta$ is logically equivalent to
$\pogformula \doteq \{\{r\}\} \cup \; \bigcup_{\nodeu \in P} \theta_{u}$
(\lstinline{equivalent st.clauseDb (r ∧ st.pogDefsForm)}), and from the invariants we can prove
$\inputformula \iff \phi_\noder$ (\lstinline{equivalent st.inputCnf (st.pog.toPropForm r)}).
After certifying a proof, the checker can pass its in-memory POG representation to the ring evaluator.
% JA: I cut a few words here to save a line.

@@

\vspace{1em}\noindent
\textbf{Model counting.} Finally, we established that ring evaluation with the appropriate weights
corresponds to the standard model count. To do so, we defined a function that
% JA: deleted the word "directly" to save a line
carries out an integer calculation of the number of models over a set of variables
of cardinality \lstinline{nVars}:
\begin{lstlisting}
def countModels (nVars : Nat) : PropForm Var → Nat
  | tr       => 2^nVars
  | fls      => 0
  | var _    => 2^(nVars - 1)
  | neg φ    => 2^nVars - countModels nVars φ
  | disj φ ψ => countModels nVars φ + countModels nVars ψ
  | conj φ ψ => countModels nVars φ * countModels nVars ψ / 2^nVars
\end{lstlisting}
We then formally proved that for a partitioned formula whose variables are among a finite set
\lstinline{s}, this computation really does count the number of models over \lstinline{s}:
\begin{lstlisting}
theorem countModels_eq_card_models {φ : PropForm Var} {s : Finset Var} :
  vars φ ⊆ s → partitioned φ → countModels (card s) φ = card (models φ s)
\end{lstlisting}
In particular, taking \lstinline{s} to be exactly the variables of \lstinline{φ},
we have that the number of models on its variables is \lstinline{countModels φ (card (vars φ))}.

>>>>

The invariants refer to the checker state \lstinline{st} with fields \lstinline{st.inputCnf} for $\inputformula$, \lstinline{st.clauseDb} for $\theta$, \lstinline{st.pog} for $P$, \lstinline{st.pogDefsForm} for the POG definitions formula $\bigwedge_{\nodeu\in P}\theta_u$, and \lstinline{st.allVars} for all variables (original and extension) added so far. For any $\nodeu\in P$, \lstinline{st.pog.toPropForm u} computes $\phi_\nodeu$. The first two invariants state that assignments to original variables extend uniquely to extension variables defining the POG nodes. In the formalization, we split this into extension and uniqueness:
\begin{lstlisting}
/-- Any assignment satisfying φ₁ extends to φ₂ while preserving values on X. -/
def extendsOver (X : Set Var) (φ₁ φ₂ : PropForm Var) :=
  ∀ (σ₁ : PropAssignment Var), σ₁ ⊨ φ₁ → ∃ σ₂, σ₁.agreeOn X σ₂ ∧ σ₂ ⊨ φ₂

@@

invariants.extends_pogDefsForm : extendsOver st.inputCnf.vars ⊤ st.pogDefsForm
invariants.uep_pogDefsForm : uniqueExt st.inputCnf.vars st.allVars st.pogDefsForm
\end{lstlisting}
The next one guarantees that the set of solutions over the original variables is preserved:
\begin{lstlisting}
def equivalentOver (X : Set Var) (φ₁ φ₂ : PropForm Var) :=
  extendsOver X φ₁ φ₂ ∧ extendsOver X φ₂ φ₁

invariants.equivInput : equivalentOver st.inputCnf.vars st.inputCnf st.clauseDb
\end{lstlisting}
Finally, for every node $\nodeu\in P$ with corresponding literal $u$ we ensure that $\phi_\nodeu$ is partitioned (\cref{def:partitioned-operation-formula}) and relate $\phi_\nodeu$ to its clausal encoding $\theta_u \doteq u \wedge \bigwedge_{\nodev\in P}\theta_v$:
\begin{lstlisting}
def partitioned : PropForm Var → Prop
  | tr | fls | var _ => True
  | neg φ    => φ.partitioned
  | disj φ ψ => φ.partitioned ∧ ψ.partitioned ∧ ∀ τ, ¬(τ ⊨ φ ∧ τ ⊨ ψ)
  | conj φ ψ => φ.partitioned ∧ ψ.partitioned ∧ φ.vars ∩ ψ.vars = ∅

invariants.partitioned : ∀ u : ILit, (st.pog.toPropForm u).partitioned
invariants.equivalent_lits : ∀ u : ILit, equivalentOver st.inputCnf.vars
    (u ∧ st.pogDefsForm) (st.pog.toPropForm x)
\end{lstlisting}
These are maintained by valid CRAT proofs. Together with a few additional invariants
that ensure the correctness of cached computations, they imply the soundness theorem for $P$ with root
node $\noder$: the \textsc{final conditions} ensure that $\theta$ is logically equivalent to
$\pogformula \doteq \{\{r\}\} \cup \; \bigcup_{\nodeu \in P} \theta_{u}$
(\lstinline{equivalent st.clauseDb (r ∧ st.pogDefsForm)}), and from the invariants we can prove
$\inputformula \iff \phi_\noder$ (\lstinline{equivalent st.inputCnf (st.pog.toPropForm r)}).
After certifying a proof, the checker can pass its in-memory POG representation to the ring evaluator.
% JA: I cut a few words here to save a line.

@@

\vspace{1em}\noindent
\textbf{Model counting.} Finally, we established that ring evaluation with the appropriate weights
corresponds to the standard model count. To do so, we defined a function that
% JA: deleted the word "directly" to save a line
carries out an integer calculation of the number of models over a set of variables
of cardinality \lstinline{nVars}:
\begin{lstlisting}
def countModels (nVars : Nat) : PropForm Var → Nat
  | tr       => 2^nVars
  | fls      => 0
  | var _    => 2^(nVars - 1)
  | neg φ    => 2^nVars - countModels nVars φ
  | disj φ ψ => countModels nVars φ + countModels nVars ψ
  | conj φ ψ => countModels nVars φ * countModels nVars ψ / 2^nVars
\end{lstlisting}
We then formally proved that for a partitioned formula whose variables are among a finite set
\lstinline{s}, this computation really does count the number of models over \lstinline{s}:
\begin{lstlisting}
theorem countModels_eq_card_models {φ : PropForm Var} {s : Finset Var} :
  vars φ ⊆ s → partitioned φ → countModels (card s) φ = card (models φ s)
\end{lstlisting}
In particular, taking \lstinline{s} to be exactly the variables of \lstinline{φ},
we have that the number of models of \lstinline{φ} on its variables is
\lstinline{countModels φ (card (vars φ))}.
